NumPy---Array Programming in Python

### 2020-02-10 body.tex
Feedback from Paul and Ross

Two Python array packages existed before NumPy.
The Numeric package began in the mid-1990s and provided an array object and array-aware functions
in Python, written in C, and linking to standard fast implementations of linear
algebra \cite{Numericmanual}.
To handle large astronomical images coming from the Hubble Space Telescope,
a reimplementation of Numeric, called Numarray, added
support for structured arrays, flexible indexing, memory mapping, byte-order variants,
more efficient memory use, flexible IEEE error handling capabilities, and
better type casting rules \cite{greenfield2003numarray}.
While Numarray was highly compatible with Numeric, the two packages
had enough differences that it divided the community, until
2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
and Numarray \cite{oliphant2006guide}---combining Numarray's features with Numeric's
performance on small arrays and its rich C \emph{Application Programming Interface} (API).

@@

@@

SciPy and Matplotlib are tightly coupled with NumPy---in terms of
history, development, and use.
SciPy provides fundamental algorithms for scientific computing,
including mathematical, scientific, and engineering routines.
Matplotlib generates publication-ready figures and visualizations.
The combination of NumPy, SciPy, and Matplotlib, together with
an advanced interactive environment like IPython \cite{perez2007ipython},
or Jupyter \cite{Kluyver:2016aa}, provides a solid foundation for array
programming in Python.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this foundation to provide several, widely used \emph{technique-specific}
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
that in turn underlay numerous \emph{domain-specific} projects
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
NumPy, as the focal point for the ecosystem of array-aware libraries,
sets documentation standards, and provides array testing infrastructure
and build support for integrating with Fortran code and using specialized
compilers and computing platforms.

@@

% FIXME SJW: Redraw the icons in vector format
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights (in blue) some of the essential tools and libraries that depend
   on NumPy.  Since most of the ecosystem
   builds on NumPy arrays and its rich and powerful collection of array-aware
   functionality, the simplest way for new array implementations to become
   available to the vast scientific Python ecosystem is by using the NumPy API and array
   protocols (Fig.~\ref{fig:array-protocol}).
  }
  \label{fig:ecosystem}
\end{figure}

Many research groups have designed large,
complex scientific libraries, which add specialized functionality for
solving problems in specific domains.
For example, the \code{eht-imaging} library \cite{chael2019ehtim} developed by the Event Horizon Telescope 
collaboration for radio interferometry imaging, analysis, and simulation
relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image
reconstruction.
SciPy is used to provide tools used in general image processing such as
filtering and image alignment, while scikit-image, an image processing
library that extends SciPy, provides higher-level functionality such as
edge filters and Hough transforms.
Other components of SciPy are used throughout the library, such as the
\code{scipy.optimize} module for handling mathematical optimization tasks.
NetworkX \cite{SciPyProceedings_11}, a package for complex
network analysis, is used for the analysis of consistency among image
comparisons.
The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
is used throughout \code{eht-imaging}, primarily to handle standard
astronomical file formats and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including for the generation of the final image of the black hole.

The rapid recent growth of data science, machine learning, and
artificial intelligence has further and dramatically boosted the usage of
scientific Python.  Examples of its significant application, such as the
above, now exist in almost every discipline in the natural and social
sciences.  These tools have become \emph{the primary}
software environment in many fields.  NumPy and its ecosystem are commonly
taught in university courses, boot camps, and summer schools, and are
at the focus of community conferences and workshops worldwide.

@@

These user-developers frequently had to write code from scratch to solve
their or their colleagues' problems---often in low-level languages
that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
To them, the advantages of an interactive, high-level array library
were evident. The design of this new tool was informed by their
experiences with powerful interactive programming languages for
scientific computing such as APL \cite{iverson1962programming} and
Yorick \cite{munro1995using}, as well as commercial languages and
environments like IDL and {MATLAB}.

@@

NumPy combines the expressive power of \emph{array programming}, 
the performance of C, and
the readability, usability, and versatility of Python in a mature,
well-tested, well-documented, and community-developed library.
It builds on decades of work refining the array programming
syntax from APL through Yorick and others.
It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.

>>>>

Two Python array packages existed before NumPy.
The Numeric package began in the mid-1990s and provided an array object and array-aware functions
in Python, written in C, and linking to standard fast implementations of linear
algebra \cite{dubois1996numerical,Numericmanual}.
To handle large astronomical images coming from the Hubble Space Telescope,
a reimplementation of Numeric, called Numarray, added
support for structured arrays, flexible indexing, memory mapping, byte-order variants,
more efficient memory use, flexible IEEE error handling capabilities, and
better type casting rules \cite{greenfield2003numarray}.
While Numarray was highly compatible with Numeric, the two packages
had enough differences that it divided the community, until
2005, when NumPy emerged as a ``best of both worlds'' unification of Numeric
and Numarray \cite{oliphant2006guide}---combining Numarray's features with Numeric's
performance on small arrays and its rich C \emph{Application Programming Interface} (API).

@@

% FIXME SJW: Redraw the icons in vector format
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{static/sketches/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights (in blue) some of the essential tools, projects, and libraries that depend
   on NumPy.  Since most of the ecosystem
   builds on NumPy arrays and its rich and powerful collection of array-aware
   functionality, the simplest way for new array implementations to become
   available to the vast scientific Python ecosystem is by using the NumPy API and array
   protocols (Fig.~\ref{fig:array-protocol}).
  }
  \label{fig:ecosystem}
\end{figure}

@@

SciPy and Matplotlib are tightly coupled with NumPy---in terms of
history, development, and use.
SciPy provides fundamental algorithms for scientific computing,
including mathematical, scientific, and engineering routines.
Matplotlib generates publication-ready figures and visualizations.
The combination of NumPy, SciPy, and Matplotlib, together with
an advanced interactive environment like IPython \cite{perez2007ipython},
or Jupyter \cite{Kluyver:2016aa}, provides a solid foundation for array
programming in Python.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this foundation to provide several, widely used \emph{technique specific}
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
that in turn underlay numerous \emph{domain specific} projects
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
NumPy, as the focal point for the ecosystem of array-aware libraries,
sets documentation standards, and provides array testing infrastructure
and build support for integrating with Fortran code and using specialized
compilers and computing platforms.

Many research groups have designed large,
complex scientific libraries, which add \emph{application specific} functionality
to the ecosystem.
For example, the \code{eht-imaging} library \cite{chael2019ehtim} developed by
the Event Horizon Telescope collaboration for radio interferometry imaging,
analysis, and simulation relies on many lower-level components of the scientific Python
ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image
reconstruction.
SciPy is used to provide tools used in general image processing such as
filtering and image alignment, while scikit-image, an image processing
library that extends SciPy, provides higher-level functionality such as
edge filters and Hough transforms.
Other components of SciPy are used throughout the library, such as the
\code{scipy.optimize} module for handling mathematical optimization tasks.
NetworkX \cite{SciPyProceedings_11}, a package for complex
network analysis, is used for the analysis of consistency among image
comparisons.
The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
is used throughout \code{eht-imaging}, primarily to handle standard
astronomical file formats and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including for the generation of the final image of the black hole.

@@

The rapid recent growth of data science, machine learning, and
artificial intelligence has further and dramatically boosted the usage of
scientific Python.  Examples of its significant application, such as the
\code{eht-imaging} library, now exist in almost every discipline in the natural and social
sciences.  These tools have become \emph{the primary}
software environment in many fields.  NumPy and its ecosystem are commonly
taught in university courses, boot camps, and summer schools, and are
at the focus of community conferences and workshops worldwide.

@@

These user-developers frequently had to write code from scratch to solve
their or their colleagues' problems---often in low-level languages
that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
To them, the advantages of an interactive, high-level array library
were evident. The design of this new tool was informed by other
powerful interactive programming languages for scientific computing
such as Basis \cite{dubois1989basis}, Yorick \cite{munro1995using}, R \cite{ihaka1996r},
and APL \cite{iverson1962programming},
as well as commercial languages and environments like IDL and {MATLAB}.

@@

NumPy combines the expressive power of \emph{array programming}, 
the performance of C, and
the readability, usability, and versatility of Python in a mature,
well-tested, well-documented, and community-developed library.
It builds on decades of work refining the array programming
syntax from APL through Basis, Yorick, and others.
It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.
